---
output:
  pdf_document: default
  html_document: default
---
Reproducible Research: Peer Assessment 2
==========================================
Jennifer Dimaano

## Impact of Severe Weather Events on Public Health and Economy in the United States


### Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


### Data Processing

#### Libraries

```{r, echo=TRUE}
library(ggplot2)
library(dplyr)
library(knitr)
library(markdown)
library(rmarkdown)
library(lattice)
knitr::opts_chunk$set(error = TRUE)

```

#### Loading Data

```{r, echo=TRUE}
if(!exists("storm.data")) {
    storm.data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"),header = TRUE)
  }
```

Data Structure

```{r, echo=TRUE}
dim(storm.data)
str(storm.data)
```

Variables that will be used:

EVTYPE: Event Type (Tornados, Flood, ….)

FATALITIES: Number of Fatalities

INJURIES: Number of Injuries

PROGDMG: Property Damage

PROPDMGEXP: Units for Property Damage (magnitudes - K,B,M)

CROPDMG: Crop Damage

CROPDMGEXP: Units for Crop Damage (magnitudes - K,BM,B)

```{r, echo=TRUE}
var <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm.data <- storm.data[var]
dim(storm.data)
```
checking data
```{r, echo=TRUE}
head(storm.data)
tail(storm.data)
# NAs in variables
sum(is.na(storm.data$EVTYPE))
sum(is.na(storm.data$FATALITIES))
sum(is.na(storm.data$INJURIES))
sum(is.na(storm.data$PROGDMG))
sum(is.na(storm.data$PROPDMGEXP))
sum(is.na(storm.data$CROPDMG))
sum(is.na(storm.data$CROPDMGEXP))
# checking dollar amounts 
sort(table(storm.data$PROPDMGEXP), decreasing = TRUE)[1:10]
sort(table(storm.data$CROPDMGEXP), decreasing = TRUE)[1:10]
# checking top 10 events
sort(table(storm.data$EVTYPE), decreasing = TRUE)[1:10]

```

Transforming variables

Group rest of values into the top 10 events by using keyword association. Events not associated with the top 10 keywords will be recorede as "OTHER." New variable EVENT is will hold the transformed records
```{r, echo=TRUE}
# create a new variable EVENT to transform variable EVTYPE in groups
storm.data$EVENT <- "OTHER"
# group by keyword in EVTYPE
storm.data$EVENT[grep("HAIL", storm.data$EVTYPE, ignore.case = TRUE)] <- "HAIL"
storm.data$EVENT[grep("HEAT", storm.data$EVTYPE, ignore.case = TRUE)] <- "HEAT"
storm.data$EVENT[grep("FLOOD", storm.data$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
storm.data$EVENT[grep("WIND", storm.data$EVTYPE, ignore.case = TRUE)] <- "WIND"
storm.data$EVENT[grep("STORM", storm.data$EVTYPE, ignore.case = TRUE)] <- "STORM"
storm.data$EVENT[grep("SNOW", storm.data$EVTYPE, ignore.case = TRUE)] <- "SNOW"
storm.data$EVENT[grep("TORNADO", storm.data$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
storm.data$EVENT[grep("WINTER", storm.data$EVTYPE, ignore.case = TRUE)] <- "WINTER"
storm.data$EVENT[grep("RAIN", storm.data$EVTYPE, ignore.case = TRUE)] <- "RAIN"
# listing the transformed event types 
sort(table(storm.data$EVENT), decreasing = TRUE)
```

Units need to be transformed. Below are the dollar conventions from [NOAA](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf):
* K or k: thousand dollars (10^3)
* M or m: million dollars (10^6)
* B or b: billion dollars (10^9)
* the rest would be consider as dollars

New variable(s) is product of value of damage and dollar unit.
```{r, echo=TRUE}
storm.data$PROPDMGEXP <- as.character(storm.data$PROPDMGEXP)
storm.data$PROPDMGEXP[is.na(storm.data$PROPDMGEXP)] <- 0 # NA's considered as dollars
storm.data$PROPDMGEXP[!grepl("K|M|B", storm.data$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
storm.data$PROPDMGEXP[grep("K", storm.data$PROPDMGEXP, ignore.case = TRUE)] <- "3"
storm.data$PROPDMGEXP[grep("M", storm.data$PROPDMGEXP, ignore.case = TRUE)] <- "6"
storm.data$PROPDMGEXP[grep("B", storm.data$PROPDMGEXP, ignore.case = TRUE)] <- "9"
storm.data$PROPDMGEXP <- as.numeric(as.character(storm.data$PROPDMGEXP))
storm.data$property.damage <- storm.data$PROPDMG * 10^storm.data$PROPDMGEXP

storm.data$CROPDMGEXP <- as.character(storm.data$CROPDMGEXP)
storm.data$CROPDMGEXP[is.na(storm.data$CROPDMGEXP)] <- 0 # NA's considered as dollars
storm.data$CROPDMGEXP[!grepl("K|M|B", storm.data$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
storm.data$CROPDMGEXP[grep("K", storm.data$CROPDMGEXP, ignore.case = TRUE)] <- "3"
storm.data$CROPDMGEXP[grep("M", storm.data$CROPDMGEXP, ignore.case = TRUE)] <- "6"
storm.data$CROPDMGEXP[grep("B", storm.data$CROPDMGEXP, ignore.case = TRUE)] <- "9"
storm.data$CROPDMGEXP <- as.numeric(as.character(storm.data$CROPDMGEXP))
storm.data$crop.damage <- storm.data$CROPDMG * 10^storm.data$CROPDMGEXP
```

### Analysis

Economic impact analysis
```{r, echo=TRUE}
storm.data.damage <- storm.data %>%
  select(EVENT, property.damage,crop.damage) %>%
  group_by(EVENT) %>%
  summarise(total.dmg = sum(property.damage,crop.damage)) %>%
  arrange(-total.dmg)
head(storm.data.damage)

```

Public health impact analysis
```{r, echo=TRUE}
# Fatalities
storm.data.fatalities <- storm.data %>% 
  select(EVENT, FATALITIES) %>% 
  group_by(EVENT) %>% 
  summarise(total.fatalities = sum(FATALITIES)) %>%
  arrange(-total.fatalities)
head(storm.data.fatalities, 10)

# Injuries
storm.data.injuries <- storm.data %>% select(EVENT, INJURIES) %>% group_by(EVENT) %>% summarise(total.injuries = sum(INJURIES)) %>% arrange(-total.injuries)
head(storm.data.injuries, 10)
```

### Results
```{r, echo=TRUE}
Fatalities <- ggplot(storm.data.fatalities, aes(x=reorder(EVENT, -total.fatalities), y=total.fatalities)) +
  geom_bar(stat = "identity") +
  xlab("Event") + 
  ylab("Total Fatalities") +
  ggtitle("Top 10 Severe Storm Events - Fatalities") +
  theme(plot.title = element_text(hjust = 0.5))

Injuries <- ggplot(storm.data.injuries, aes(x=reorder(EVENT, -total.injuries), y=total.injuries)) +
  geom_bar(stat = "identity") +
  xlab("Event") + 
  ylab("Total Injuries") +
  ggtitle("Top 10 Severe Storm Events - Injuries") +
  theme(plot.title = element_text(hjust = 0.5))

Damages <- ggplot(storm.data.damage, aes(x=reorder(EVENT, -total.dmg), y=(total.dmg/1000000000))) +
  geom_bar(stat = "identity") +
  xlab("Event") + 
  ylab("Total Damages ($ in Billions)") +
  ggtitle("Top Severe Storm Events with the largest Economic Impact") +
  theme(plot.title = element_text(hjust = 0.5))

print(Fatalities)
print(Injuries)
print(Damages)
```

Tornados make up the largest public health impact resulting in 91,407 annual injuries and 5,661 annual fatalities. Heat and other severe storm events follow after Tornados. Hail makes up the least amount of fatalities at 15, while rain produces only 305 injuries over a year.

In terms of economic impact, flood damaage resulted in over \$179B. While Tornados made up a fraction of that damage at \$5B.
